Disclosing funding sources for open access publication fees: the Open APC

### 2016-05-31 article.Rmd
back-uping

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. However, collecting information about what was funded is in most cases difficult. One reason why payments made for open access journal publications are often hard to track is that they are fragmented across the budgets of funding agencies, research institutions, and libraries, or, if no institutional support exists, are taken from personal budgets. Institutional open access funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets stand in contrast with those support structures and are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Along with the fragmentation of payments, another problem is that funding for open access publications lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014]. This situation is similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to a complex landscape of variable pricing schemes [@Pinfield_2015] include submission or page charges [@BJ_RK_2012].

@@

This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Research addressing the question about how much was spent on open access journal articles either surveyed authors [@soap], collected price information from journal websites [@Walters_2011, @Solomon_2012], or used self-reported data [@Pinfield_2015]. In the latter case, some research funders have begun collecting and disclosing expenditures for open access journal articles as open data.

To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditures along with bibliographic information, including title, journal and publisher, persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc).

In examining self-reported institutional payments made for open access articles in Germany, this paper seeks to contribute to the evolving empirical evidence on questions about open access publication fees as business model in academic publishing. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has been encouraging open access publishing since years. The DFG launched its "Open-Access Publishing" program in 2009. This programme has strongly influenced how university libraries support open access publication fees through funds.[^3] To reduce administrative burdens, grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers. Grantees are also required to report the institutional publication output and their publication fee spending to the DFG on a regular basis, and to present the university-wide strategy to sustain the funds when DFG’s initial support runs out lately in 2019.

@@

Instead of re-using publisher and journal information contained in the Open APC data-set, we fetched bibliographic metadata for each article from CrossRef on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from CrossRef as well, this information was retrieved at the time when the participating institutions submitted the cost information. The Open APC initiative tracked the date of data submission and the contributors for every data submission with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, CrossRef regularly updated metadata to represent ongoing mergers of publishing houses or name changes. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from CrossRef for the whole Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access CrossRef's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Since CrossRef was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection in this study also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we reviewed the openly available spreadsheets from the Wellcome Trust [@wellcome_apc], the Austrian Science Fund FWF [@fwf_apc] and Jisc. For analysis, we obtained the amount of overall spending on publications fees, as well as on charges to publish in fully open access journal articles. In the case of FWF, we gathered the cost information from accompanying reports. We used the spreadsheet data to summarize  Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average Euro foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded Jisc's spending in 2014 because full currency information in the data-set was not available.

Data collection methods of the Open APC initiative and those of the others differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institution were required to report, publication identifier in the Wellcome Trust data was also added through automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided only to compare German spending with that of other initiatives, but not its size and distribution over publisher and journal titles, as well as the indexing coverage in CrossRef. However, we made use of the literature to discuss our findings in relation to these Austrian and UK initiatives.

>>>>

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. Yet, how and to what extent these activities are effective in terms of the number of supported research articles and associated costs remains under debate. One reason why studying what was funded is in most cases difficult is that payments are fragmented across the budgets of funding agencies, research institutions, and libraries, or, if  institutional support is limited, are taken from personal budgets. Recent research indicate that institutional open access funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets stand in contrast with those support structures and are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Along with the fragmentation of payments, another problem in the empirical study of open access journals is that funding for open access publications lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to a complex landscape of variable pricing schemes [@Pinfield_2015] include submission or page charges [@BJ_RK_2012].

@@

This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Research addressing the question about how much was spent on open access journal articles either surveyed authors [@soap], collected price information from journal websites [@Walters_2011; @Solomon_2012], or used self-reported data [@Pinfield_2015].

-> summary of empirical findings

<!-- In the latter case, some research funders have begun collecting and disclosing expenditures for open access journal articles as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditures along with bibliographic information, including title, journal and publisher, persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc). -->

In examining self-reported institutional payments made for open access articles in Germany, this paper seeks to contribute to the evolving empirical evidence on questions about open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has been encouraging open access publishing since years. The DFG launched its "Open-Access Publishing" program in 2009. This programme has strongly influenced how university libraries support open access publication fees through funds.[^3] To reduce administrative burdens, grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers. Grantees are also required to report the institutional publication output and their publication fee spending to the DFG on a regular basis, and to present the university-wide strategy to sustain the funds when DFG’s initial support runs out lately in 2019.

@@

We fetched bibliographic metadata for each article from CrossRef on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from CrossRef as well, this information was retrieved at the time when the participating institutions submitted the cost information. The Open APC initiative tracked the date of data submission and the contributors for every data submission with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, CrossRef regularly updated metadata to represent ongoing mergers of publishing houses or name changes. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from CrossRef for the whole Open APC data-set instead of re-using publisher and journal information contained in the Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access CrossRef's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because CrossRef was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection in this study also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we reviewed the openly available data-sets from the Wellcome Trust [@wellcome_apc], the Austrian Science Fund FWF [@fwf_apc] and Jisc. For analysis, we obtained the amount of overall spending on publications fees, as well as on charges to publish in fully open access journal articles. In the case of FWF, we gathered the cost information from accompanying reports. We used the spreadsheet data to summarize  Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average Euro foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded Jisc's spending in 2014 because full currency information in the data-set was not available.

Data collection methods of the Open APC initiative and those of the others differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institution were required to report, publication identifier in the Wellcome Trust data was also added through automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided only to compare German spending with that of other initiatives, but not its size and distribution over publisher and journal titles, as well as the indexing coverage in CrossRef.
